Formalization of Typed Lua

### 2015-06-12 doc/papers/paper/paper.tex
DLS paper: reformatting

In the {\tt then} branch it is guaranteed that both {\tt q}
and {\tt r} are integers, and in the {\bf else} branch it is
guaranteed that {\tt q} is {\tt nil} and {\tt r} is a string.
In Typed Lua, the return type of {\tt idiv} can be a union of
tuples describing the possible return values. Typed Lua deconstructs unions of tuples in a way that the dependencies between the types of each value in the tuple can be tracked.

@@

\begin{figure}
\label{fig:subtuples}
\[
\begin{array}{c}

\]
\caption{Subtyping for second-level types}	
\end{figure}

Figure~\ref{fig:subtuples} shows the subtyping rules for
second-level types. 
Tuple types are covariant, so the subtyping rule for pair types is straightforward:

%\[
%\begin{array}{c}
%\mylabel{S-PAIR}\\
%\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
%       \senv \vdash P_{1} \subtype P_{2}}
%      {\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
%\end{array}
%\]

@@

The remaining subtyping rules for second-level types
reflect this semantics. Rule {\sc S-VARIADIC} states
that variadic types are covariant (just having $F_1{*} \subtype F_2{*}$ iff $F_1 \subtype F_2$ would be too restrictive,
as $\Nil{*}$ would not be a subtype of $\Number{*}$):
\[
\begin{array}{c}
\mylabel{S-VARIADIC}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \cup \Nil}
      {\senv \vdash F_{1}{*} \subtype F_{2}{*}}
\end{array}
\]

Rules {\sc S-VAR-EXPAND} and {\sc S-VAR-DROP} state the
conditions when variadic tuples of different length are
compatible:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-VAR-EXPAND}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \;\;\;
       \senv \vdash F_{1}{*} \subtype P_{2}}
      {\senv \vdash F_{1}{*} \subtype F_{2} \times P_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VAR-DROP}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \cup \Nil \;\;\;
       \senv \vdash P_{1} \subtype F_{2}{*}}
      {\senv \vdash F_{1} \times P_{1} \subtype F_{2}{*}}
\end{array}
\end{array}
\]

@@

We will now focus on the subtyping relation among table
types. Subtyping between a {\em fixed} or {\em closed}
table type and a {\em closed} table type resembles the
standard subtyping rule between mutable records, with
width subtyping plus depth subtyping from a mutable
or immutable field to an immutable field:
\[
\begin{array}{c}
\mylabel{S-TABLE-FC-C}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..n \; \exists j \in 1..m \\
       \senv \vdash K_{j} \subtype K_{i}' \;\;\;
       \senv \vdash K_{i}' \subtype K_{j} \;\;\;
       \senv \vdash V_{j} \subtype_{c} V_{i}'
       \end{array}}
      {\begin{array}{c}
       \senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{fixed|closed} \subtype\\
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}
       \end{array}} \; m \ge n
\end{array}
\]

Rule \textsc{S-TABLE-FC-C} uses the auxiliary relation $\subtype_{c}$ for handling depth subtyping:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD-C-MM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2} \;\;\;
       \senv \vdash V_{2} \subtype V_{1}}
      {\senv \vdash V_{1} \subtype_{c} V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD-C-II}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash \Const \; V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD-C-MI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}
\end{array}
\]

The subtyping relation between {\em unique} and {\em closed}
table types is more relaxed, as the typing rules of Section~\ref{sec:tables} guarantee that a {\em unique}
table type can only appear on the left side of a subtyping
judgment if the reference to the table with the {\em unique}
tag is being destroyed:
\[
\begin{array}{c}
\mylabel{S-TABLE-U-C}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \; \forall j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
\end{array}
\]

Rule {\sc S-TABLE-U-C} allows with subtyping, depth subtyping
on any field, not only immutable fields, as well as the addition
of new fields and the merging of existing fields, as long as the types of the values of new and merged fields include $\Nil$.
This rule uses the auxiliary relations $\subtype_{u}$ and $\subtype_{o}$, which respectively allow depth subtyping on all fields, and the addition of new fields:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD-U-MM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash V_{1} \subtype_{u} V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD-U-IM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash \Const \; V_{1} \subtype_{u} V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD-U-II}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash \Const \; V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD-U-MI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD-O-M}\\
\dfrac{\senv \vdash \Nil \subtype V}
      {\senv \vdash \Nil \subtype_{o} V}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD-O-I}\\
\dfrac{\senv \vdash \Nil \subtype V}
      {\senv \vdash \Nil \subtype_{o} \Const \; V}
\end{array}
\end{array}
\]

The rationale for rule {\sc S-TABLE-U-C} derives from the
fact that reading a non-existing key just yield $\Nil$
instead of being an error in Lua semantics, plus the fact
that the type system guarantees that a table with a
{\em unique} type has no keys with a value that does not
belong to one of its key types, and does not have any
aliases. A table with a type $\{1{:}\String, 2{:}\String,
``x"{:}\Integer, ``y"{:}\Value\}_{unique}$ can be safely recast as a
table with type 
$\{\Number{:}\String \cup \Nil, ``x"{:}\Number, ``z"{:}
\Number \cup \Nil\}_{closed}$ as long as it cannot be
accessed anymore with the original type.

Subtyping from an {\em open} table type to a {\em closed}
table type also allows width subtyping, merging fields,
and adding new fields, but forbids depth subtyping between
mutable fields or an immutable and a mutable field, as
the typing rules guarantee that the {\em open} reference
is destroyed, but the {\em open} table may already have {\em closed} aliases:
\[
\begin{array}{c}
\mylabel{S-TABLE-O-C}\\
\dfrac{\begin{array}{c}
	\forall i \in 1..m \; \forall j \in 1..n \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \in 1..n \; \not\exists i \in 1..m \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
\end{array}
\]

The subtyping relation from {\em unique} to {\em unique},
{\em open} or {\em fixed} table types is still flexible, but forbids standard width subtyping, as all these table types keep the guarantee that the values of all keys in tables of these types belong to one of their key types. Merging fields and adding new fields are still allowed:
\[
\begin{array}{c}
\mylabel{S-TABLE-U-UOF}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\begin{array}{c}
       \senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype\\
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{unique|open|fixed}
       \end{array}}
\end{array}
\]

The subtyping relation between {\em open} and {\em open} or
{\em fixed} table types mixes the depth subtyping restrictions of rule {\sc S-TABLE-O-C} with the width subtyping restrictions of
rule {\sc S-TABLE-U-UOF}, but still allow merging fields and
adding new fields:
\[
\begin{array}{c}
\mylabel{S-TABLE-O-OF}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\begin{array}{c}
       \senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype\\
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{open|fixed}
       \end{array}}
\end{array}
\]

Finally, the subtyping relation between {\em fixed} table
types is the more restrictive, and does not allow width
subtyping, merging fields, or adding fields, just standard
depth subtyping:
\[
\begin{array}{c}
\mylabel{S-TABLE-F-F}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..n \; \exists j \in 1..n \\
       \senv \vdash K_{j} \subtype K_{i}' \;\;\;
       \senv \vdash K_{i}' \subtype K_{j} \;\;\;
       \senv \vdash V_{j} \subtype_{c} V_{i}'
       \end{array}}
      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{fixed} \subtype
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{fixed}}
\end{array}
\]

@@

Subtyping between two recursive types uses the \emph{Amber rule}~\cite{cardelli1986amber}:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-AMBER}\\
\dfrac{\senv[x_{1} \subtype x_{2}] \vdash T_{1} \subtype T_{2}}
      {\senv \vdash \mu x_{1}.T_{1} \subtype \mu x_{2}.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-ASSUMPTION}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
      {\senv \vdash x_{1} \subtype x_{2}}
\end{array}
\end{array}
\]

Subtyping between a recursive type and a non-recursive type, or
vice-versa, uses a pair of rules that unfold the recursive type:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-UNFOLD-R}\\
\dfrac{\senv \vdash T_{1} \subtype [x \mapsto \mu x.T_{2}]T_{2}}
      {\senv \vdash T_{1} \subtype \mu x.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNFOLD-L}\\
\dfrac{\senv \vdash [x \mapsto \mu x.T_{1}]T_{1} \subtype T_{2}}
      {\senv \vdash \mu x.T_{1} \subtype T_{2}}
\end{array}
\end{array}
\]

@@

\begin{figure*}[!ht]
	\textbf{Abstract Syntax}\\
	\dstart
	$$
	\begin{array}{rlr}
	s ::= & & \textsc{statements:}\\
	& \;\; \mathbf{skip} & \textit{skip}\\
	& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
	& | \; \overline{l} = el & \textit{multiple assignment}\\
	& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
	| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
	& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s & \textit{recursive function} \\
	& | \; \mathbf{return} \; el & \textit{return} \\
	& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
	& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
	e ::= & & \textsc{expressions:}\\
	& \;\; \mathbf{nil} & \textit{nil}\\
	& | \; k & \textit{other literals}\\
	& | \; id & \textit{variable access}\\
	& | \; e_{1}[e_{2}] & \textit{table access}\\
	& | \; {<}T{>} \; id & \textit{type coercion}\\
	& | \; f & \textit{function declaration}\\
	& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
	| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
	& | \; e_{1} + e_{2} \;
	| \; e_{1} \; {..} \; e_{2} \;
	| \; e_{1} == e_{2} \;
	| \; e_{1} < e_{2} & \textit{binary operations}\\
	& | \; e_{1} \;\&\; e_{2} \;
	| \; e_{1} \; \mathbf{and} \; e_{2} \;
	| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
	& | \; \mathbf{not} \; e \;
	| \; \# \; e & \textit{unary operations} \\
	& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
	l ::= & & \textsc{left-hand values:}\\
	& \;\; id_{l} & \textit{variable assignment}\\
	& | \; e_{1}[e_{2}] & \textit{table assignment}\\
	& | \; id[k] \; {<}T{>} & \textit{type coercion}\\
	k ::= & & \textsc{literal constants:}\\
	& \;\; \mathbf{false} \; | \;
	\mathbf{true} \; | \;
	{\it int} \; | \;
	{\it float} \; | \;
	{\it string} & \\
	el ::= & & \textsc{expression lists:}\\
	& \;\; \overline{e} \; | \;
	\overline{e}, me & \\
	me ::= & & \textsc{multiple results:}\\
	& \;\; a & \textit{application}\\
	& | \; {...} & \textit{vararg expression}\\
	a ::= & & \textsc{applications:}\\
	& \;\; e(el) & \textit{function application}\\
	& | \; e{:}n(el) & \textit{method application}\\
	f ::= & & \textsc{function declarations:}\\
	& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
	pl ::= & & \textsc{parameter lists:}\\
	& \;\; \overline{id{:}T} \; | \;
	\overline{id{:}T},{...}{:}T & \\
	\end{array}
	$$
	\dend
	\caption{The abstract syntax of Core Typed Lua}
	\label{fig:syntax}
\end{figure*}

@@

As an example, consider again the integer division function
from Section~1, which returns the quotient and rest of integer
division, or a division by zero error, now statically typed:
\begin{verbatim}
  local function idiv(x : integer, y : integer):
        (integer, integer) | (nil, string)
    if divisor == 0 then
      return nil, "division by zero"
    else
      return x // y, x % y
    end 
  end
\end{verbatim}

>>>>

In the {\bf then} branch it is guaranteed that both {\tt q}
and {\tt r} are integers, and in the {\bf else} branch it is
guaranteed that {\tt q} is {\tt nil} and {\tt r} is a string.
In Typed Lua, the return type of {\tt idiv} can be a union of
tuples describing the possible return values. Typed Lua deconstructs unions of tuples in a way that the dependencies between the types of each value in the tuple can be tracked.

@@

\begin{figure}[t]
\label{fig:subtuples}
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-Pair}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
	\senv \vdash P_{1} \subtype P_{2}}
{\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
\end{array}
\begin{array}{c}
\mylabel{S-Variadic}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \cup \Nil}
{\senv \vdash F_{1}{*} \subtype F_{2}{*}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-VarExpand}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \;\;\;
	\senv \vdash F_{1}{*} \subtype P_{2}}
{\senv \vdash F_{1}{*} \subtype F_{2} \times P_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VarDrop}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \cup \Nil \;\;\;
	\senv \vdash P_{1} \subtype F_{2}{*}}
{\senv \vdash F_{1} \times P_{1} \subtype F_{2}{*}}
\end{array}
\end{array}
\]
\caption{Subtyping for second-level types}	
\end{figure}

Figure~\ref{fig:subtuples} shows the subtyping rules for
second-level types. 
Tuple types are covariant, so the subtyping rule {\sc S-Pair} for pair types  is straightforward.

@@

The remaining subtyping rules for second-level types
reflect this semantics. Rule {\sc S-Variadic} states
that variadic types are covariant (just having $F_1{*} \subtype F_2{*}$ iff $F_1 \subtype F_2$ would be too restrictive,
as $\Nil{*}$ would not be a subtype of $\Number{*}$).
Rules {\sc S-VarExpand} and {\sc S-VarDrop} state the
conditions when variadic tuples of different length are
compatible.

@@

\begin{figure}[t]
\label{fig:subtable}
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TableFCtoC}\\
\dfrac{\begin{array}{c}
	\forall i \in 1..n \; \exists j \in 1..m \\
	\senv \vdash K_{j} \subtype K_{i}' \;\;\;
	\senv \vdash K_{i}' \subtype K_{j} \;\;\;
	\senv \vdash V_{j} \subtype_{c} V_{i}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{fixed|closed} \subtype\\
	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}
	\end{array}} \; m \ge n
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableUtoC}\\
\dfrac{\begin{array}{c}
	\forall i \in 1..m \; \forall j \in 1..n \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
	\forall j \in 1..n \; \not\exists i \in 1..m \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype
	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableOtoC}\\
\dfrac{\begin{array}{c}
	\forall i \in 1..m \; \forall j \in 1..n \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \in 1..n \; \not\exists i \in 1..m \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableUtoUOF}\\
\dfrac{\begin{array}{c}
	\forall i \in 1..m \\
	\exists j \in 1..n \;
	\senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
	\forall j \in 1..n \; \not\exists i \in 1..m \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype\\
	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{unique|open|fixed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableOtoOF}\\
\dfrac{\begin{array}{c}
	\forall i \in 1..m \\
	\exists j \in 1..n \;
	\senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \in 1..n \; \not\exists i \in 1..m \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype\\
	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{open|fixed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableFtoF}\\
\dfrac{\begin{array}{c}
	\forall i \in 1..n \; \exists j \in 1..n \\
	\senv \vdash K_{j} \subtype K_{i}' \;\;\;
	\senv \vdash K_{i}' \subtype K_{j} \;\;\;
	\senv \vdash V_{j} \subtype_{c} V_{i}'
	\end{array}}
{\senv \vdash \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{fixed} \subtype
	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{fixed}}
\end{array}
\end{array}
\]
\caption{Subtyping for table types}
\end{figure}

\begin{figure}[t]
\label{fig:subfield}
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FieldClsMtoM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2} \;\;\;
	\senv \vdash V_{2} \subtype V_{1}}
{\senv \vdash V_{1} \subtype_{c} V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldClsItoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash \Const \; V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldClsMtoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-FieldUnqMtoM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash V_{1} \subtype_{u} V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldUnqItoM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash \Const \; V_{1} \subtype_{u} V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldUnqItoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash \Const \; V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldUnqMtoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldOpnNtoM}\\
\dfrac{\senv \vdash \Nil \subtype V}
{\senv \vdash \Nil \subtype_{o} V}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldOpnNtoI}\\
\dfrac{\senv \vdash \Nil \subtype V}
{\senv \vdash \Nil \subtype_{o} \Const \; V}
\end{array}
\end{array}
\]
\caption{Subtyping for fields}
\end{figure}

Figure~\ref{fig:subtable} shows the subtyping rules for
table types. These rules use the auxiliary relations
$\subtype_{c}$, $\subtype_{u}$, and $\subtype_{o}$
shown in Figure~\ref{fig:subfield} to handle depth
subtyping where applicable.

Rule {\sc S-TableFCtoC} for subtyping
between a {\em fixed} or {\em closed}
table type and a {\em closed} table type resembles the
standard subtyping rule between mutable records, with
width subtyping plus depth subtyping from a mutable
or immutable field to an immutable field.

Rule {\sc S-TableUtoC} for subtyping between {\em unique} and {\em closed} table types is more permissive, as the typing rules of Section~\ref{sec:tables} guarantee that a {\em unique}
table type can only appear on the left side of a subtyping
judgment if the reference to the table with the {\em unique}
tag is being destroyed. This rule allows with subtyping, depth subtyping on any field, not only immutable fields, as well as the addition of new fields and the merging of existing fields, as long as the types of the values of new and merged fields include $\Nil$.

The rationale for the permissiveness of rule {\sc S-TableUtoC} derives from the fact that reading a non-existing key just yields $\Nil$ instead of being an error in Lua semantics, plus the fact
that the type system guarantees that a table with a
{\em unique} type has no keys with a value that does not
belong to one of its key types, and does not have any
aliases. With these guarantess, a table with type $\{1{:}\String, 2{:}\String,
``x"{:}\Integer, ``y"{:}\Value\}_{unique}$ can be safely recast as a table with type 
$\{\Number{:}\String \cup \Nil, ``x"{:}\Number, ``z"{:}
\Number \cup \Nil\}_{closed}$ as long as it becomes
inaccessible with the original type, which Section~\ref{sec:tables} will show that is the case
for any type judgement where this subtyping rule is used.

Rule {\sc S-TableOtoC} for subtyping from an {\em open} table type to a {\em closed} table type also allows width subtyping, merging fields, and adding new fields, but forbids depth subtyping between mutable fields or an immutable and a mutable field, as the typing rules guarantee that the {\em open} reference
is destroyed, but the {\em open} table may already have {\em closed} aliases.

Rule {\sc S-TableUtoUOF} for subtyping from {\em unique} to {\em unique}, {\em open} or {\em fixed} table types is still permissive, but forbids standard width subtyping, as all these table types keep the guarantee that the values of all keys in tables of these types belong to one of their key types. Merging fields and adding new fields are still allowed.

Rule {\sc S-TableOtoOF} for subtyping from {\em open} to {\em open} or {\em fixed} table types mixes the depth subtyping restrictions of rule {\sc S-TableOtoC} with the width subtyping restrictions of rule {\sc S-TableUtoUOF}, but still allows merging fields and adding new fields.

Finally, rule {\sc S-TableFtoF} for subtyping between {\em fixed} table types is the most restrictive, and does not allow width
subtyping, merging fields, or adding fields, just standard
depth subtyping.

@@

\begin{figure}[t]
\label{fig:subrec}
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-Amber}\\
\dfrac{\senv[x_{1} \subtype x_{2}] \vdash T_{1} \subtype T_{2}}
{\senv \vdash \mu x_{1}.T_{1} \subtype \mu x_{2}.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-Assumption}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
{\senv \vdash x_{1} \subtype x_{2}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-UnfoldRight}\\
\dfrac{\senv \vdash T_{1} \subtype [x \mapsto \mu x.T_{2}]T_{2}}
{\senv \vdash T_{1} \subtype \mu x.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UnfoldLeft}\\
\dfrac{\senv \vdash [x \mapsto \mu x.T_{1}]T_{1} \subtype T_{2}}
{\senv \vdash \mu x.T_{1} \subtype T_{2}}
\end{array}
\end{array}
\]
\caption{Subtyping for recursive types}
\end{figure}

Figure~\ref{fig:subrec} shows the subtyping rules for
recursive types. Rule {\sc S-Amber} is the \emph{Amber rule}~\cite{cardelli1986amber} for subtyping between
two recursive types. Rules {\sc S-UnfoldRight} and {\sc S-UnfoldLeft} unfold the recursive type for establishing
subtyping between a non-recursive and a recursive type,
or vice-versa.

@@

\begin{figure*}[!ht]
	$$
	\begin{array}{rlr}
	s ::= & & \textsc{statements:}\\
	& \;\; \mathbf{skip} & \textit{skip}\\
	& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
	& | \; \overline{l} = el & \textit{multiple assignment}\\
	& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
	| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
	& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s & \textit{recursive function} \\
	& | \; \mathbf{return} \; el & \textit{return} \\
	& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
	& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
	e ::= & & \textsc{expressions:}\\
	& \;\; \mathbf{nil} & \textit{nil}\\
	& | \; k & \textit{other literals}\\
	& | \; id & \textit{variable access}\\
	& | \; e_{1}[e_{2}] & \textit{table access}\\
	& | \; {<}T{>} \; id & \textit{type coercion}\\
	& | \; f & \textit{function declaration}\\
	& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
	| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
	& | \; e_{1} + e_{2} \;
	| \; e_{1} \; {..} \; e_{2} \;
	| \; e_{1} == e_{2} \;
	| \; e_{1} < e_{2} & \textit{binary operations}\\
	& | \; e_{1} \;\&\; e_{2} \;
	| \; e_{1} \; \mathbf{and} \; e_{2} \;
	| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
	& | \; \mathbf{not} \; e \;
	| \; \# \; e & \textit{unary operations} \\
	& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
	l ::= & & \textsc{left-hand values:}\\
	& \;\; id_{l} & \textit{variable assignment}\\
	& | \; e_{1}[e_{2}] & \textit{table assignment}\\
	& | \; id[k] \; {<}T{>} & \textit{type coercion}\\
	k ::= & & \textsc{literal constants:}\\
	& \;\; \mathbf{false} \; | \;
	\mathbf{true} \; | \;
	{\it int} \; | \;
	{\it float} \; | \;
	{\it string} & \\
	el ::= & & \textsc{expression lists:}\\
	& \;\; \overline{e} \; | \;
	\overline{e}, me & \\
	me ::= & & \textsc{multiple results:}\\
	& \;\; a & \textit{application}\\
	& | \; {...} & \textit{vararg expression}\\
	a ::= & & \textsc{applications:}\\
	& \;\; e(el) & \textit{function application}\\
	& | \; e{:}n(el) & \textit{method application}\\
	f ::= & & \textsc{function declarations:}\\
	& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
	pl ::= & & \textsc{parameter lists:}\\
	& \;\; \overline{id{:}T} \; | \;
	\overline{id{:}T},{...}{:}T & \\
	\end{array}
	$$
	\caption{Abstract Syntax}
	\label{fig:syntax}
\end{figure*}

@@

As an example, consider again the {\tt idiv} function
from Figure~\ref{fig:over}, which returns the quotient and rest of integer division, or a division by zero error. In Typed Lua, we can give a return type $(\Integer \times \Integer \times \Nil*) \sqcup (\Nil \times \String \times \Nil{*})$ to that function.
